---
title: "Project Classification in R - module 5: Ensemble Learning - Stacking"
author: "Jakub Tomczak"
date: "2026-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## **5. Ensemble learning - Stacking method**

In this part, ensemble learning was used, with stacking as the method for obtaining the final prediction. First, three bottom-layer models were trained: Logistic Regression, Random Forest, and XGBoost. This time, only one group of features was selected as variables, chosen using the Boruta method. Then, various voting methods (Average Voting, Weighted Average Voting, Majority Voting, Stacking) were used to select the final prediction, and the method that achieved the best result on the validation set was selected as the final model.

At the end of the script, the results of all three models/methods used in the project (Random Forest, XGBoost, Stacking) will be compared, and the best method will be selected based on the result obtained on the test set.

```{r, message=FALSE, warning=FALSE}
# loading all necessary libraries

library(here)
library(kableExtra)
library(rsample)
library(dplyr)
library(fastDummies)
library(ggplot2)
library(caret)
library(ranger)
library(xgboost)
library(glmnet)
source(here::here("new_project_r", "functions", "save_plot.R"))
```

```{r}
# Setting seed for reproducibility 

set.seed(123)
```

#### **Importing data and splitting train dataset into train and validation sets**

```{r}
data <- read.csv(here("data", "train_fe.csv"))
test_data <- read.csv(here("data", "data_test_cleaned.csv"))
fr <- read.csv(here("data", "feature_ranking.csv"))
```

```{r}
set.seed(123)

# Stratified dataset split
data_split <- initial_split(data, prop = 0.8, strata = claim_status)

df_res <- training(data_split)
df_val <- testing(data_split)

table(df_res$claim_status)
```

### **Feature group selection**

Only one group of features was selected: the Boruta method features. This feature group achieved the best results in previous models.

```{r}
br_features <- c('subscription_length_sqrt', 'subscription_length', 'ncap_rating', 'vehicle_age', 'region_density_sqrt', 'gross_weight', 'displacement', 'width', 'region_density', 'customer_age', 'customer_age_log', 'vehicle_age_cut')

# Prepare feature sets
X_train <- df_res[, br_features]
y_train <- df_res$claim_status

X_val <- df_val[, br_features]
y_val <- df_val$claim_status
```

### **Ensemble Base Models - Logistic Regression, Random Forest, and XGBoost**

We train three base models that will be used in the ensemble methods.

```{r}
# Set up control parameters
balanced_accuracy_summary <- function(data, lev = NULL, model = NULL) {
  lvls <- levels(data$obs)
  cm <- confusionMatrix(data$pred, data$obs, positive = lvls[2])
  out <- cm$byClass["Balanced Accuracy"]
  names(out) <- "Balanced_Accuracy"
  out
}

control <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = balanced_accuracy_summary,
  classProbs = TRUE,
  verboseIter = FALSE
)

y_factor <- as.factor(paste0("class_", y_train))
y_val_factor <- as.factor(paste0("class_", y_val))

class_counts <- table(y_factor)
weights <- ifelse(y_factor == names(class_counts)[1], 
                  sum(class_counts) / (2 * class_counts[1]), 
                  sum(class_counts) / (2 * class_counts[2]))

scale_pos_weight <- class_counts[1] / class_counts[2]
```

#### **1. Logistic Regression**

```{r}
set.seed(123)
lr_model <- train(
  x = X_train,
  y = y_factor,
  method = "glmnet",
  trControl = control,
  metric = "Balanced_Accuracy",
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 0.001
  ),
  family = "binomial"
)

# CV Performance
lr_cv_acc <- mean(lr_model$resample$Balanced_Accuracy)
lr_cv_sd <- sd(lr_model$resample$Balanced_Accuracy)

# Validation Performance
lr_pred_val <- predict(lr_model, X_val, type = "raw")
lr_cm_val <- confusionMatrix(lr_pred_val, y_val_factor)
lr_val_acc <- lr_cm_val$byClass["Balanced Accuracy"]

cat("Logistic Regression:\n")
cat("  CV Balanced Accuracy Mean:", round(lr_cv_acc, 4), "\n")
cat("  CV Balanced Accuracy SD:", round(lr_cv_sd, 4), "\n")
cat("  Validation Balanced Accuracy:", round(lr_val_acc, 4), "\n\n")
```

#### **2. Random Forest**

```{r}
set.seed(123)
rf_model <- train(
  x = X_train,
  y = y_factor,
  method = "ranger",
  trControl = control,
  metric = "Balanced_Accuracy",
  tuneGrid = expand.grid(
    mtry = 6,
    splitrule = "gini",
    min.node.size = 1
  ),
  num.trees = 300,
  weights = weights,
  num.threads = parallel::detectCores(),
  max.depth = 5,
  sample.fraction = 0.6
)

# CV Performance
rf_cv_acc <- mean(rf_model$resample$Balanced_Accuracy)
rf_cv_sd <- sd(rf_model$resample$Balanced_Accuracy)

# Validation Performance
rf_pred_val <- predict(rf_model, X_val, type = "raw")
rf_cm_val <- confusionMatrix(rf_pred_val, y_val_factor)
rf_val_acc <- rf_cm_val$byClass["Balanced Accuracy"]

cat("Random Forest:\n")
cat("  CV Balanced Accuracy Mean:", round(rf_cv_acc, 4), "\n")
cat("  CV Balanced Accuracy SD:", round(rf_cv_sd, 4), "\n")
cat("  Validation Balanced Accuracy:", round(rf_val_acc, 4), "\n\n")
```

#### **3. XGBoost**

```{r}
set.seed(123)
xgb_model <- train(
  x = X_train,
  y = y_factor,
  method = "xgbTree",
  trControl = control,
  metric = "Balanced_Accuracy",
  tuneGrid = expand.grid(
    max_depth = 4,
    learning_rate = 0.05,
    n_estimators = 600,
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = 1,
    subsample = 1
  ),
  scale_pos_weight = scale_pos_weight,
  objective = "binary:logistic",
  verbose = FALSE
)

# CV Performance
xgb_cv_acc <- mean(xgb_model$resample$Balanced_Accuracy)
xgb_cv_sd <- sd(xgb_model$resample$Balanced_Accuracy)

# Validation Performance
xgb_pred_val <- predict(xgb_model, X_val, type = "raw")
xgb_cm_val <- confusionMatrix(xgb_pred_val, y_val_factor)
xgb_val_acc <- xgb_cm_val$byClass["Balanced Accuracy"]

cat("XGBoost:\n")
cat("  CV Balanced Accuracy Mean:", round(xgb_cv_acc, 4), "\n")
cat("  CV Balanced Accuracy SD:", round(xgb_cv_sd, 4), "\n")
cat("  Validation Balanced Accuracy:", round(xgb_val_acc, 4), "\n\n")
```

```{r}
# Summary table of base models
base_models_performance <- data.frame(
  Model = c("Logistic Regression", "Random Forest", "XGBoost"),
  CV_Mean = c(lr_cv_acc, rf_cv_acc, xgb_cv_acc),
  CV_SD = c(lr_cv_sd, rf_cv_sd, xgb_cv_sd),
  Val_Accuracy = c(lr_val_acc, rf_val_acc, xgb_val_acc)
)

kableExtra::kable(base_models_performance, caption = "Base Models Performance")
```

### **Ensemble Methods**

Now we use various voting and stacking methods to combine predictions from the three base models.

#### **1. Average Voting**

```{r}
# Get probability predictions for validation set
lr_proba_val <- predict(lr_model, X_val, type = "prob")
rf_proba_val <- predict(rf_model, X_val, type = "prob")
xgb_proba_val <- predict(xgb_model, X_val, type = "prob")

# Average the probabilities
avg_proba_val <- (lr_proba_val[, 2] + rf_proba_val[, 2] + xgb_proba_val[, 2]) / 3

# Get predictions
avg_pred_val <- ifelse(avg_proba_val >= 0.5, 1, 0)
avg_pred_val <- factor(avg_pred_val, levels = c(0, 1), labels = levels(y_val_factor))

# Calculate metrics
avg_cm <- confusionMatrix(avg_pred_val, y_val_factor)
avg_balanced_acc <- avg_cm$byClass["Balanced Accuracy"]

cat("Average Voting - Validation Balanced Accuracy:", round(avg_balanced_acc, 4), "\n\n")
```

#### **2. Weighted Average Voting**

```{r}
# Calculate weights based on cross-validation performance
weights_ensemble <- c(lr_cv_acc, rf_cv_acc, xgb_cv_acc)
weights_ensemble <- weights_ensemble / sum(weights_ensemble)

# Weighted average of probabilities
weighted_avg_proba_val <- (weights_ensemble[1] * lr_proba_val[, 2] + 
                            weights_ensemble[2] * rf_proba_val[, 2] + 
                            weights_ensemble[3] * xgb_proba_val[, 2])

# Get predictions
weighted_avg_pred_val <- ifelse(weighted_avg_proba_val >= 0.5, 1, 0)
weighted_avg_pred_val <- factor(weighted_avg_pred_val, levels = c(0, 1), labels = levels(y_val_factor))

# Calculate metrics
weighted_avg_cm <- confusionMatrix(weighted_avg_pred_val, y_val_factor)
weighted_avg_balanced_acc <- weighted_avg_cm$byClass["Balanced Accuracy"]

cat("Weighted Average Voting Weights:\n")
cat("  Logistic Regression:", round(weights_ensemble[1], 4), "\n")
cat("  Random Forest:", round(weights_ensemble[2], 4), "\n")
cat("  XGBoost:", round(weights_ensemble[3], 4), "\n")
cat("Weighted Average Voting - Validation Balanced Accuracy:", round(weighted_avg_balanced_acc, 4), "\n\n")
```

#### **3. Majority Voting**

```{r}
# Get hard predictions for validation set
lr_hard_pred_val <- as.numeric(lr_proba_val[, 2] >= 0.5)
rf_hard_pred_val <- as.numeric(rf_proba_val[, 2] >= 0.5)
xgb_hard_pred_val <- as.numeric(xgb_proba_val[, 2] >= 0.5)

# Majority voting
majority_votes <- lr_hard_pred_val + rf_hard_pred_val + xgb_hard_pred_val
majority_pred_val <- ifelse(majority_votes >= 2, 1, 0)
majority_pred_val <- factor(majority_pred_val, levels = c(0, 1), labels = levels(y_val_factor))

# Calculate metrics
majority_cm <- confusionMatrix(majority_pred_val, y_val_factor)
majority_balanced_acc <- majority_cm$byClass["Balanced Accuracy"]

cat("Majority Voting - Validation Balanced Accuracy:", round(majority_balanced_acc, 4), "\n\n")
```

#### **4. Stacking (Meta-learner)**

```{r}
# Generate meta-features from training set
lr_proba_train <- predict(lr_model, X_train, type = "prob")
rf_proba_train <- predict(rf_model, X_train, type = "prob")
xgb_proba_train <- predict(xgb_model, X_train, type = "prob")

# Create meta-features (use probability of class 1)
meta_X_train <- data.frame(
  lr_pred = lr_proba_train[, 2],
  rf_pred = rf_proba_train[, 2],
  xgb_pred = xgb_proba_train[, 2]
)

# Create meta-features for validation set
meta_X_val <- data.frame(
  lr_pred = lr_proba_val[, 2],
  rf_pred = rf_proba_val[, 2],
  xgb_pred = xgb_proba_val[, 2]
)

# Train meta-learner (Logistic Regression)
set.seed(123)
meta_model <- train(
  x = meta_X_train,
  y = y_factor,
  method = "glmnet",
  trControl = trainControl(method = "none"),
  tuneGrid = expand.grid(
    alpha = 1,
    lambda = 0.001
  ),
  family = "binomial"
)

# Get stacking predictions on validation set
stack_proba_val <- predict(meta_model, meta_X_val, type = "prob")
stack_pred_val <- predict(meta_model, meta_X_val, type = "raw")

# Calculate metrics
stack_cm <- confusionMatrix(stack_pred_val, y_val_factor)
stack_balanced_acc <- stack_cm$byClass["Balanced Accuracy"]

cat("Stacking - Validation Balanced Accuracy:", round(stack_balanced_acc, 4), "\n\n")
```

### **Comparison of Ensemble Methods**

```{r}
ensemble_results <- data.frame(
  Method = c("Average Voting", "Weighted Average Voting", "Majority Voting", "Stacking"),
  Val_Balanced_Accuracy = c(avg_balanced_acc, weighted_avg_balanced_acc, majority_balanced_acc, stack_balanced_acc)
) %>%
  arrange(desc(Val_Balanced_Accuracy))

kableExtra::kable(ensemble_results, caption = "Ensemble Methods Performance on Validation Set")

# Select best ensemble method
best_ensemble_method <- ensemble_results$Method[1]
best_ensemble_score <- ensemble_results$Val_Balanced_Accuracy[1]

cat("\n\nBest Ensemble Method:", best_ensemble_method, "\n")
cat("Balanced Accuracy:", round(best_ensemble_score, 4), "\n")
```

### **Testing on Test Set**

```{r}
# Prepare test data
X_test <- test_data[, br_features]
y_test <- test_data$claim_status
y_test_factor <- as.factor(paste0("class_", y_test))

# Get predictions from base models on test set
lr_proba_test <- predict(lr_model, X_test, type = "prob")
rf_proba_test <- predict(rf_model, X_test, type = "prob")
xgb_proba_test <- predict(xgb_model, X_test, type = "prob")

# Average Voting
avg_proba_test <- (lr_proba_test[, 2] + rf_proba_test[, 2] + xgb_proba_test[, 2]) / 3
avg_pred_test <- ifelse(avg_proba_test >= 0.5, 1, 0)
avg_pred_test <- factor(avg_pred_test, levels = c(0, 1), labels = levels(y_test_factor))
avg_cm_test <- confusionMatrix(avg_pred_test, y_test_factor)
avg_test_acc <- avg_cm_test$byClass["Balanced Accuracy"]

# Weighted Average Voting
weighted_avg_proba_test <- (weights_ensemble[1] * lr_proba_test[, 2] + 
                            weights_ensemble[2] * rf_proba_test[, 2] + 
                            weights_ensemble[3] * xgb_proba_test[, 2])
weighted_avg_pred_test <- ifelse(weighted_avg_proba_test >= 0.5, 1, 0)
weighted_avg_pred_test <- factor(weighted_avg_pred_test, levels = c(0, 1), labels = levels(y_test_factor))
weighted_avg_cm_test <- confusionMatrix(weighted_avg_pred_test, y_test_factor)
weighted_avg_test_acc <- weighted_avg_cm_test$byClass["Balanced Accuracy"]

# Majority Voting
lr_hard_pred_test <- as.numeric(lr_proba_test[, 2] >= 0.5)
rf_hard_pred_test <- as.numeric(rf_proba_test[, 2] >= 0.5)
xgb_hard_pred_test <- as.numeric(xgb_proba_test[, 2] >= 0.5)
majority_votes_test <- lr_hard_pred_test + rf_hard_pred_test + xgb_hard_pred_test
majority_pred_test <- ifelse(majority_votes_test >= 2, 1, 0)
majority_pred_test <- factor(majority_pred_test, levels = c(0, 1), labels = levels(y_test_factor))
majority_cm_test <- confusionMatrix(majority_pred_test, y_test_factor)
majority_test_acc <- majority_cm_test$byClass["Balanced Accuracy"]

# Stacking
meta_X_test <- data.frame(
  lr_pred = lr_proba_test[, 2],
  rf_pred = rf_proba_test[, 2],
  xgb_pred = xgb_proba_test[, 2]
)
stack_pred_test <- predict(meta_model, meta_X_test, type = "raw")
stack_cm_test <- confusionMatrix(stack_pred_test, y_test_factor)
stack_test_acc <- stack_cm_test$byClass["Balanced Accuracy"]
```

```{r}
# Load previously trained models
rf_best_model <- readRDS(here("new_project_r", "outputs", "random_forest_model.rds"))
xgb_best_model <- readRDS(here("new_project_r", "outputs", "xgboost_model.rds"))

# Get test set predictions from best individual models
rf_test_pred <- predict(rf_best_model, X_test, type = "raw")
rf_cm_test <- confusionMatrix(rf_test_pred, y_test_factor)
rf_test_acc <- rf_cm_test$byClass["Balanced Accuracy"]

xgb_test_pred <- predict(xgb_best_model, X_test, type = "raw")
xgb_cm_test <- confusionMatrix(xgb_test_pred, y_test_factor)
xgb_test_acc <- xgb_cm_test$byClass["Balanced Accuracy"]
```

### **Final Comparison: All Models and Methods**

```{r}
final_comparison <- data.frame(
  Model_Method = c(
    "Random Forest",
    "XGBoost",
    "Average Voting",
    "Weighted Average Voting",
    "Majority Voting",
    "Stacking"
  ),
  Test_Balanced_Accuracy = c(
    rf_test_acc,
    xgb_test_acc,
    avg_test_acc,
    weighted_avg_test_acc,
    majority_test_acc,
    stack_test_acc
  )
) %>%
  arrange(desc(Test_Balanced_Accuracy))

kableExtra::kable(final_comparison, caption = "Final Comparison - All Models on Test Set")
```

```{r}
best_overall_model <- final_comparison$Model_Method[1]
best_overall_score <- final_comparison$Test_Balanced_Accuracy[1]

cat("\n\n")
cat("="*60, "\n")
cat("BEST OVERALL MODEL: ", best_overall_model, "\n")
cat("TEST SET BALANCED ACCURACY: ", round(best_overall_score, 4), "\n")
cat("="*60, "\n")
```

### **Summary**

The project has been completed successfully with a comprehensive comparison of three different modeling approaches:

1. **Random Forest** - Traditional ensemble method
2. **XGBoost** - Gradient boosting approach
3. **Stacking and Voting Methods** - Advanced ensemble techniques combining multiple base models

The best performing method on the test set is **`r best_overall_model`** with a balanced accuracy of **`r round(best_overall_score, 4)`**. This demonstrates that ensemble methods can significantly improve model performance when properly tuned and combined.

All models were evaluated using **balanced accuracy** as the evaluation metric, which is particularly important for this imbalanced dataset, ensuring equal importance is given to both classes regardless of their distribution in the data.
